Data Analysis Methods for the Detection of the Epoch of Reionization

### 2017-07-06 PAPER.tex
finished section 2.1

\begin{abstract}
The Epoch of Reionization (EoR) is an uncharted era in our Universe's history during which the birth of the first stars and galaxies led to the ionization of neutral hydrogen. This important epoch of our cosmic dawn harbors a wealth of information regarding the environment during this transformative time, including insight into the nature of the first luminous sources and implications about cosmological parameters. There are many experiments investigating the EoR by tracing the $21$ cm line of neutral hydrogen, a signal which is very faint and difficult to isolate. With a new generation of instruments and a statistical power spectrum detection in our foreseeable future, it has become increasingly important to develop techniques that help maximize sensitivity and validate results. Additionally, it is imperative to understand the trade-offs between different methods and their effects on common power spectrum themes. In this paper, we focus on three major themes - signal loss, power spectrum error bar estimation, and bias in measurements. We describe techniques that affect these themes using both a toy model and data taken by the 64-element configuration of the Donald C. Baker Precision Array for Probing the Epoch of Reionization (PAPER).
\end{abstract}

@@

By about one billion years after the Big Bang, the very first stars and galaxies are thought to have ionized all the neutral hydrogen that dominated the Universe's early life. This important transition, during which the first luminous structures formed from gravitational collapse and emit intense radiation, transforming the cold neutral gas into a plasma, is known as the epoch of reionization (EoR). The EoR represents an unexplored era in our cosmic dawn, whose history encodes important information regarding the nature of the first galaxies and the process of structure formation. A direct detection of the EoR would unlock powerful information about the intergalactic medium, revealing connections between the smooth matter distribution exhibited via cosmic microwave background (CMB) studies and the highly structured web of galaxies we observe today.

One promising technique to probe the EoR is to target the 21 cm wavelength emission that is emitted by neutral hydrogen via its spin-flip transition. This technique is powerful because it can be observed as a function of redshift - that is, the wavelength of the signal reaching our telescopes can be directly mapped to a distance from where the emission originated before stretching out as it traveled through expanding space. The 21 cm line therefore offers a window into following the evolution of ionization, temperature, and density fluctuations on cosmic scales.

Although a detection of the EoR has not currently been made to-date, there are several radio telescope experiments that have succeeded in using the 21 cm signal from hydrogen in order to place constraints on the brightness of the EoR. Examples of experiments investigating the mean brightness temperature of the EoR relative to the CMB are EDGES (\citealt{bowman2010}), the LWA (\citealt{ellingson_et_al2009}), LEDA (\citealt{greenhill_bernardi2012}), DARE (\citealt{burns2012}), SciHi (\citealt{voytek2014}), BIGHORNS (\citealt{sokolowski2015}), and SARAS (\citealt{patra2015}). Major interferometers, which seek to measure statistical power spectra, include the GMRT (\citealt{paciga_et_al2013}), LOFAR (\citealt{van_haarlem_et_al2013}), the MWA (\citealt{tingay_et_al2013}), the 21CMA (\citealt{peterson_et_al2004}, \citealt{wu2009}), and PAPER (\citealt{parsons_et_al2010}). The Hydrogen Epoch of Reionization Array (HERA), which is currently being built, is a next-generation instrument that hopes to combine lessons learned from previous experiments and is forecasted to be able to make a successful \cc{?$\sigma$} detection with an eventual \cc{?} elements.

The major challenge that faces all 21 cm experiments is in isolating a small signal that is buried underneath foregrounds and instrumental systematics that are 4-5 orders of magnitude brighter. A clean measurement therefore requires an intimate understanding of the instrument and a very careful, thorough understanding of data analysis choices. With HERA on the horizon and continual progress being made in the field, it is becoming increasingly more important to understand how the methods we choose affect each other and any power spectrum results. In this paper, we discuss three themes essential to a $21$ cm power spectrum study and how data analysis choices affect each. We approach these themes from both a broad perspective, and through a detailed case study using data from the 64-element configuration of PAPER.

@@

Going back to our time-averaged simulation data, we will now describe several ways to alter the covariance matrix $\textbf{C}$ in order to avoid signal loss. Recall that $\textbf{C}$ is computed from the dataset itself, leading to the up-weighting of very noisy, non-existent modes. In practice we compute $\textbf{C}$ this way because we do not know how to accurately model the matrix. But for our toy model, we know that our simulated, perfectly-noisy EoR should have a covariance matrix that mimics the identity matrix, with its variance encoded along the diagonal. If we model $\textbf{C}_{EoR}$ as such and add it to $\textbf{C}_{FG} = \textbf{x}_{FG}\textbf{x}_{FG}^{\dagger}$ to obtain a final $\textbf{C}$ to use in weighting (instead of computing it based on $(\textbf{x}_{FG} + \textbf{x}_{EoR})$), we see that there is no signal loss (Figure \ref{fig:toy_sigloss8}, left).

\begin{figure*}
	\centering
	\includegraphics[height=0.21\textwidth]{plots/toy_sigloss10.png}
	\includegraphics[height=0.21\textwidth]{plots/toy_sigloss8.png}
	\includegraphics[height=0.21\textwidth]{plots/toy_sigloss9.png}
	\caption{Power spectra of foregrounds only (blue), EoR only (red), and the weighted combined dataset (green). We show three alternate weighting options that each avoid signal loss, including modeling the covariance matrix of EoR (left), regularizing $\textbf{C}$ by adding an identity matrix to it (middle), and using only the first few eigenmodes of $\textbf{C}$ (right).}
	\label{fig:toy_sigloss8}
\end{figure*}

\cc{add paragraph about regularization options}

@@

The Donald C. Baker Precision Array for Probing the Epoch of Reionization (PAPER) is a dedicated 21 cm experiment located in the Karoo Desert in South Africa. The PAPER-64 configuration consists of 64 dual-polarization drift-scan elements, each 2 m on a side. The antenna layout is formatted in a grid layout (Figure \ref{fig:paper}), with 8 antennas on a side, 30 m spacing between antennas along the East/West direction, and 4 m spacing between antennas along the North/South direction. For the rest of this paper, we focus on data from only the 30 m pure East/West baselines.

>>>>

\begin{abstract}
The Epoch of Reionization (EoR) is an uncharted era in our Universe's history during which the birth of the first stars and galaxies led to the ionization of neutral hydrogen. This important epoch of our cosmic dawn harbors a wealth of information regarding the environment during this transformative time, including insight into the nature of the first luminous sources and implications about cosmological parameters. There are many experiments investigating the EoR by tracing the $21$ cm line of neutral hydrogen, a signal which is very faint and difficult to isolate. With a new generation of instruments and a statistical power spectrum detection in our foreseeable future, it has become increasingly important to develop techniques that help maximize sensitivity and validate results. Additionally, it is imperative to understand the trade-offs between different methods and their effects on common power spectrum themes. In this paper, we focus on three major themes - signal loss, power spectrum error bar estimation, and bias in measurements. We describe techniques that affect these themes using both a toy model and data taken by the 64-element configuration of the Donald C. Backer Precision Array for Probing the Epoch of Reionization (PAPER).
\end{abstract}

@@

By about one billion years after the Big Bang, the very first stars and galaxies are thought to have ionized all the neutral hydrogen that dominated the baryonic matter content in the early Universe. This transition period, during which the first luminous structures formed from gravitational collapse and began to emit intense radiation that ionized the cold neutral gas into a plasma, is known as the epoch of reionization (EoR). The EoR is a relatively unexplored era in our cosmic dawn. Its history encodes important information regarding the nature of the first galaxies and the processes of structure formation. Direct measurements of the EoR would unlock powerful information about the intergalactic medium, revealing connections between the smooth matter distribution exhibited via cosmic microwave background (CMB) studies and the highly structured web of galaxies we observe today.

One promising technique to probe the EoR is to target the 21 cm wavelength emission that is emitted by neutral hydrogen via its spin-flip transition. This technique is powerful because it can be observed as a function of redshift --- that is, the wavelength of the signal reaching our telescopes can be directly mapped to a distance from where the emission originated before stretching out as it traveled through expanding space. The 21 cm line therefore offers a window into the evolution of ionization, temperature, and density fluctuations on cosmic scales.

Although a detection of the EoR remains elusive, there are several radio telescope experiments that have succeeded in using the 21 cm signal from hydrogen to place constraints on the brightness of the EoR. Examples of experiments investigating the mean brightness temperature of the EoR relative to the CMB are EDGES (\citealt{bowman2010}), the LWA (\citealt{ellingson_et_al2009}), LEDA (\citealt{greenhill_bernardi2012}), DARE (\citealt{burns2012}), SciHi (\citealt{voytek2014}), BIGHORNS (\citealt{sokolowski2015}), and SARAS (\citealt{patra2015}). Radio interferometers which seek to measure statistical power spectra include the GMRT (\citealt{paciga_et_al2013}), LOFAR (\citealt{van_haarlem_et_al2013}), the MWA (\citealt{tingay_et_al2013}), the 21CMA (\citealt{peterson_et_al2004}, \citealt{wu2009}), and PAPER (\citealt{parsons_et_al2010}). The Hydrogen Epoch of Reionization Array (HERA), which is currently being built, is a next-generation instrument that aims to combine lessons learned from previous experiments and is forecasted to be able to make a successful $23\sigma$ detection with an eventual $350$ elements, using current analysis techniques (\citealt{deboer_et_al2017}, \citealt{pober_et_al2014}).

The major challenge that faces all 21 cm experiments is isolating a small signal that is buried underneath foregrounds and instrumental systematics that are 4-5 orders of magnitude brighter. A clean measurement therefore requires an intimate understanding of the instrument and a rigorous study of data analysis choices. With continual progress being made in the field and HERA on the horizon, it is becoming increasingly important to understand how the methods we choose interact with each other to affect power spectrum results. In this paper, we discuss three themes essential to a $21$ cm power spectrum analysis and how data analysis choices affect each. We approach these themes from a broad perspective, and then perform a detailed case study using data from the 64-element configuration of PAPER.

@@

Going back to our time-averaged simulation data, we will now describe several ways to alter the covariance matrix $\textbf{C}$ in order to avoid signal loss. Recall that $\textbf{C}$ is computed from the dataset itself, leading to the up-weighting of very noisy, non-existent modes. In practice we compute $\textbf{C}$ this way because we do not know how to accurately model the matrix. But for our toy model, we know that our simulated, perfectly-noisy EoR should have a covariance matrix that mimics the identity matrix, with its variance encoded along the diagonal. If we model $\textbf{C}_{EoR}$ as such and add it to $\textbf{C}_{FG} = \textbf{x}_{FG}\textbf{x}_{FG}^{\dagger}$ to obtain a final $\textbf{C}$ to use in weighting (instead of computing it based on $(\textbf{x}_{FG} + \textbf{x}_{EoR})$), we see that there is no signal loss (Figure \ref{fig:toy_sigloss8}, upper left).

\begin{figure*}
	\centering
	\includegraphics[height=0.25\textwidth]{plots/toy_sigloss10.png}
	\includegraphics[height=0.25\textwidth]{plots/toy_sigloss8.png}
	\includegraphics[height=0.25\textwidth]{plots/toy_sigloss9.png}
	\includegraphics[height=0.25\textwidth]{plots/toy_sigloss11.png}
	\caption{Power spectra of foregrounds only (blue), EoR only (red), and the weighted combined dataset (green). We show four alternate weighting options that each avoid signal loss, including modeling the covariance matrix of EoR (upper left), regularizing $\textbf{C}$ by adding an identity matrix to it (upper right), using only the first few eigenmodes of $\textbf{C}$ (lower left), and multiplying an identity matrix to $\textbf{C}$ (lower right).}
	\label{fig:toy_sigloss8}
\end{figure*}

The next three panels of Figure \ref{fig:toy_sigloss8} each involve methods that regularize the matrix $\textbf{C}$ after it is computed from the foreground + EoR dataset. By regularizing $\textbf{C}$, we are flattening out its eigenspectrum. Therefore, some modes will be down-weighted, but the rest will be treated with similar weights. In other words, we won't be up-weighting the last few noisy modes.

Regularization schemes that we are highlighting here are adding an identity matrix to $\textbf{C}$ (Figure \ref{fig:toy_sigloss8}, upper right), killing all but the first three eigenmodes of $\textbf{C}$ (lower left), and multiplying an identity matrix to $\textbf{C}$ (lower right). Each avoids signal loss, but it is clear that two of the methods down-weight foregrounds much better than the third. For this toy model, our foregrounds are spread out in frequency and therefore have non-negligible frequency-frequency correlations. Therefore, when multiplying $\textbf{C}$ by an identity matrix, we are only preserving correlations between the same frequencies. Hence, because we killed off information from all other frequency combinations, we do a poor job suppressing the foreground.

Although this method did not successfully recover EoR for this particular simulation, it is important that we show that there are many options for estimating a covariance matrix, and some may work better than others based on the dataset. One may imagine a situation where a particular systematic is contained to an isolated frequency. In such a case, preserving only the diagonal elements of $\textbf{C}$ would be an effective way of removing this contamination. In Section \ref{sec:Sigloss}, we apply the lessons we've learned about signal loss and its interaction with fringe-rate filtering and inverse covariance weighting to PAPER-64 data.

@@

The Donald C. Backer Precision Array for Probing the Epoch of Reionization (PAPER) is a dedicated 21 cm experiment located in the Karoo Desert in South Africa. The PAPER-64 configuration consists of 64 dual-polarization drift-scan elements, each 2 m on a side. The antenna layout is formatted in a grid layout (Figure \ref{fig:paper}), with 8 antennas on a side, 30 m spacing between antennas along the East/West direction, and 4 m spacing between antennas along the North/South direction. For the rest of this paper, we focus on data from only the 30 m pure East/West baselines.
